Interactive Filter and Display of Hillary Clinton's Emails: \\A Cautionary Tale of Metadata

### 2017-03-31 VAST_submission/vgtc_journal_latex/paper.tex
Submission

\abstract{We present a web-based visualization that allows the user to interactively filter and display characteristics of 32,795 Hillary Clinton's emails as provided by Wikileaks.

The visualization focuses on the meta-data of each email, including its senders, receivers, and the timestamp the email appeared on the Clinton server.  An interactive time range slider filters all email and all displays automatically update to changes in the slider.  The main display shows Clinton's most frequent correspondents arranged as nodes of a spoked graph with Clinton at the centre.  Volume determines the thickness of each spoke and high volume determines an inner circle whose spokes are shortened.  Correspondents and their edges are coloured according to whether that email account could be identified as being an approved Federal government account or not.  A second display shows two daily time series: the total number of emails for that day, and the number meeting selection criteria.  A third display shows a scatterplot of the time of day versus the day on which that email appeared.  Scatterplot points are coloured by whether the email was redacted or not.

@@

The implementation is described over the next three sections, beginning with discussion of the data in Section \ref{sect:data} which discusses the source of the data, the extraction and cleaning process in Subsection \ref{sect:data:extrclean}, the details of the metadata in Subsection \ref{sect:data:metadata}, the use of redaction codes in Subsection \ref{sect:data:redact} and content data used in Subsection \ref{sect:data:contentinfo}. The displays of the data and their patterns over the entire unfiltered data are then described in Section \ref{sect:Displays}. Within this section, Subsection \ref{sect:Displays:circle} describes the spoked network plot, Subsection \ref{sect:Displays:volume} addresses the email volume time series plot, Subsection \ref{sect:Displays:times} discusses the scatterplot of daily sending times, Subsection \ref{sect:Displays:FOIA} outlines the exemption code barplot, and Subsection \ref{sect:Displays:TFIDF} explains the content tf-idf and frequency information. Interactive filters which are implemented in the application are discussed in Section \ref{sect:Filters}. Finally, brief analysis of some interesting findings is included in Section \ref{sect:Analysis}, alongside a summary of the timeline of the Clinton email imbroglio in Section \ref{sect:Timeline}.

%\section{Background}
%\label{sect:Background}
%
%\subsection{A brief timeline on the private email server}
%\label{sect:Background:Timeline}
\section{A brief timeline on the private email server}
\label{sect:Timeline}
While the following timeline is somewhat abbreviated, it should serve to raise some of the major issues and concerns related to the private email server and its contents.  It also introduces some of the principal characters involved.  More complete and in depth timelines are readily available elsewhere (e.g. \cite{attkissonTimeline, thompsonTimeline, WashPostTimeline, clintonWikipedia, TimeMagEverything}).

@@

Public knowledge that a private email server being used by Secretary Clinton and others for State Department and personal communications did not surface until March 2015 \cite{NewYorkTimes2015} during the course of a U.S. Congressional investigation \cite{BenghaziReport} of the September 11, 2012 attack by militants on U.S. compounds in Benghazi Libya.

@@

Combetta, Mills, and Samuelson will later be granted partial immunity by the Justice Department during the FBI investigations into the private email server, as were two others: Bryan Pagliano (original server manager) and  John Bentel (former director of Information Resources Management for the State Department's Executive Secretariat) \cite{immunityPolitico, immunityDailyCaller, immunityIT}.

On April 12, 2015 Clinton announces that she is running for the U.S. Presidency.  On July 24, 2015, inspectors general for the State Department  and the national intelligence agencies announce finding classified information in the emails and that the information they found was classified at the time sent \cite{serverClassified}, though her campaign declared that they must have been classified after the fact. On August 19, 2015, Clinton calls the allegation of mishandling classified information a ``disagreement between agencies'' \cite{clintonDenialGuardian}.

Nearly one year later, July 5, 2016,  FBI Director James Comey recommended that no charges be laid against Clinton on use of private email server \cite{nochargeFBI}.  On October 28, 2016, Comey revealed that in a separate investigation into former Congressman Anthony Weiner, that emails belonging to his wife Huma Abedin have been found on his laptop.  Since Abedin was a close aid to former Secretary Clinton, FBI investigations were reopened into the private server usage but closed again by Comey on November 6, 2016 without charges being laid \cite{nochargeFBINov}.  In both cases, Comey and the FBI are criticized by pundits from different political parties.

@@

\subsection{Metadata}
\label{sect:data:metadata}
For each email, from the HTML we extract as best we can the identity of the persons sending and receiving the email as well as the date and time at which the email was processed by the server. The fields used were the address to, address from, contact name to, contact name from, subject line, and time. As well, forwarding chains of email addresses were captured through the identification of any to or from fields followed by emails within the text. In cases where no contact name was present the address was substituted. When the address was missing no imputation was completed. Carbon copy information was not extracted.

The HTML was constructed from email printed out, redacted, and then provided as pdfs by the State Department.  Consequently, detailed email header information as would normally be available electronically is mostly missing.   All time stamps appear to be the local date and time at which the server sent or received that email (e.g. no time zone or other source time or IP chain information is available).  Moreover, because of redaction (typically FOIA exemption B6 ~\cite{FOIA}),  sender and receiver emails may have truncated domains,  contain only the person's name, or be missing altogether.  In cases where both \texttt{From} and \texttt{To} are entirely missing, but there is an email chain within the message, we do not impute values  (e.g. see \texttt{https://wikileaks.org/clinton-emails/emailid/31599}).

@@

\section{Displays}
\label{sect:Displays}
\TODO{Each of these four displays is described and commented on in context of whole time line}
Four displays of the metadata and quasi-metadata

\subsection{Inner circle}
\label{sect:Displays:circle} 
This display takes the most frequent correspondents (max. 20) in the selected emails and arranges their email addresses  at the ends of  equiangular spokes having Secretary Clinton as hub.  The hub is coloured red to show that this account is known not to be a government sponsored account (either \texttt{.gov} of \texttt{.mil}).  Every account that is identifiably not a government sponsored account is coloured red.  Those that are identifiably government are coloured either blue (for \texttt{.gov}) or black (for \texttt{.mil}, if any appear).  Those which can not be determined as either government or not, are coloured orange.

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{SpiralNetworkFullTime}
\caption{Inner Circle over all correspondence}
\label{fig:InnerCircleAll}
\end{center}
\end{figure}
Figure~\ref{fig:InnerCircleAll} shows the top 20 correspondents (or ``Inner Circle'')  over all emails in the collection.
Note that Huma Abedin appears both as a blue  \texttt{.gov} account and as a red non-government (\texttt{clintonemail.com}) account.   The other red account is that of Sidney Blumenthal.  This has been a source of some controversy  \cite{BlumenthalControversy, NYTBlumenthalBenghazi}, since he had been rejected by the State Department yet appears to provide Clinton advice throughout her tenure.  Wendy Sherman appears as orange. Sherman was appointed by Clinton to be Under Secretary of State for Political Affairs in 2011 and the high frequency of Sherman emails dates from this time.  Nevertheless, from this archive the email address could not be identified definitively as being either government or non-government.  Either the address was redacted for privacy reasons (a B6 FOIA exemption), or it was simply unavailable in the header provided.

The greater the number of emails between correspondents, the wider is their connecting spoke.  Whenever the difference in volume is great enough,  correspondents are separated into two groups: those with the greatest correspondence have shorter spokes visually placing them ``closer'' to Clinton.   As Figure~\ref{fig:InnerCircleAll} shows shows, Clinton's closest inner circle are her closest aides Huma Abedin, Cheryl Mills, and Jake Sullivan. \TODO{Chris, can you be more precise about the algorithm?}
\subsection{Email volume}
\label{sect:Displays:volume}
Figure~\ref{fig:VolumeAll} 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{VolumeFullTime}
\caption{Email Volume: All emails (top) ; Those sent by Clinton (bottom)}
\label{fig:VolumeAll}
\end{center}
\end{figure}
presents two time series showing the volume of emails for each day.  In grey at top is the daily total; in magenta at bottom is the total for the subset chosen by the filters.  In Figure~\ref{fig:VolumeAll}, the bottom series is the total daily emails sent from Clinton to others.

@@

\subsection{Term frequency and TFIDF}
\label{sect:Displays:TFIDF}
The 20 terms (stemmed words excluding stopwords) which appear in the greatest number of the emails selected are displayed as those having ``Highest Frequency''.  
\begin{figure}[h]
\begin{center}
\begin{tabular}{ccc}
\includegraphics[width=0.4\linewidth]{HighFreqAll} &
$~~~~$ &
\includegraphics[width=0.4\linewidth]{HighTFIDFAll} 
\end{tabular}
\caption{Top 20 terms appearing in all emails}
\label{fig:top20all}
\end{center}
\end{figure}
Those terms which appear frequently within some emails but less frequently across emails will have a high tf-idf (term frequency - inverse document frequency) score.  The top 20 scoring of these in the selected emails are shown in the ``TFIDF'' display. \TODO{Chris, check that I have described these two measures correctly}
The top 20 shown in Figure~\ref{fig:top20all} are for all emails in the corpus; these will change depending on the filtering.

@@

\subsection{The last two years}
One might begin an interactive analysis by looking at the email patterns in the last two years of Clinton's tenure as Secretary of State.  This is done by simply having the right end of the time slider of Figure~\ref{fig:timeSlider} as far right as possible and then moving the left most end towards the middle of the time period until January 1, 2011 is reached.  Figure~\ref{fig:lastTwoYears} 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{EmailVolumeB1LastTwoYears}
\caption{Last two years:  Selected Emails are those that are B1 redacted}
\label{fig:lastTwoYears}
\end{center}
\end{figure}
shows the resulting email volume time series over this period.  Note, however, that in addition to filtering on time we have chosen to also filter on FOIA exemption code, selecting only those emails containing one or more B1 (national security and foreign policy matters) redactions.  The magenta time series at the bottom then shows all emails, sent or received by Clinton, in the last two years of her tenure which were redacted in part by the State Department (after the fact, not at the time) as being B1 FOIA exempt.  As can be seen, there are relatively few of these.

Recall from Section~\ref{sect:Displays:circle} that the tallest spike of Figure~\ref{fig:lastTwoYears} corresponds to the beginning of the battle for Tripoli (August 21, 2011).   Another series of the next largest spikes appear at the far left of this time window, at the beginning of 2011.  We might choose to compress the time window more, to focus on this section of the emails.

@@

Slightly to the right of this peak, there are several more.   At this time, international forces are gathering to consider Libya and its leader Muammar Gaddafi.  Consulting Jake Sullivan's ``tick tock on Libya'' timeline describing Secretary Clinton's leadership on this, we move the sliders in to focus on the time range from February 25 when Secretary Clinton announced the suspension of the Libyan embassy in Washington, to March 18, the day after Secretary Clinton had secured ``\ldots Russian abstention and Portuguese and African support for UNSC 1973, ensuring that  it passes. 1973 authorizes a no-fly zone over Libya and `all necessary measures' - code for military action - to protect civilians against Gaddafi's army \ldots'' \cite{tickTockLibya}.  On March 19, 2011 NATO military operations in Libya began \cite{LibyaTimelineWiki}.

The readjusted time series are shown in Figure~\ref{fig:EmailVolumeB1LibyaBuildup}.
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{EmailVolumeB1LibyaBuildup}
\caption{Feb. 25 -- March 15, 2011:  Selected Emails are those that are B1 redacted}
\label{fig:EmailVolumeB1LibyaBuildup}
\end{center}
\end{figure}
The left most spike is the rightmost spike of approximately 80 emails from Figure~{EmailVolumeB1JanApril2011}.   
Of course all other displays have been updating every time the time slider is adjusted.

@@

What other restrictions, as measured by the FOIA exemption codes, can be seen in Figure~\ref{fig:FOIAB12011LibyaBuildup}.
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{FOIAB12011LibyaBuildup}
\caption{FOIA code distribution: All  B1 redacted, Feb. 25 -- March 18, 2011}
\label{fig:FOIAB12011LibyaBuildup}
\end{center}
\end{figure}
There appear to be about 100 emails in the selected set.  In addition to having B1 redactions, many of them also have B5 (inter- or intra-agency) and B6 (personal) redactions.

We might now consider the composition of the inner circle for this set of emails.  Figure~\ref{fig:InnerCircle2011B1LibyaBuildup} 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{InnerCircle2011B1LibyaBuildup}
\caption{Inner Circle: All  B1 redacted, Jan. 1 -- Apr 15, 2011}
\label{fig:InnerCircle2011B1LibyaBuildup}
\end{center}
\end{figure}
shows the set of correspondents.  Not surprisingly, Jake Sullivan is the principal correspondent.  Huma Abedin is next, though she is using two emails, one a government account the other a personal \texttt{clintonemail.com} account.  Melanie Verveer has the least amount of traffic (judging by the lightness of her spoke colour).  Sidney Blumenthal, a non-government employee, is in this inner circle of B1 exempted emails.  So too is former U.K. Prime Minister Tony Blair; this is the unidentifiable domain account \texttt{aclb}.
It should be noted that there could be other accounts which do not appear in the inner circle because their emails were not present in the email record.  See Section~\ref{sect:data:metadata}.

@@

The inner circle for all emails from September 11 to November 23, 2012 is shown in Figure~\ref{fig:teaser}(b).  Those closest to Clinton in terms of volume of correspondence are Clinton aides Jake Sullivan, Cheryl Mills, and Monica Hanley.  Monica Hanley is also an account whose identity as government or not could not be determined.  Tony Blair and Sidney Blumenthal both appear in the inner circle, as does Wendy Sherman (though the account could not be identified as government or not).  Oscar Flores is an assistant to the Clintons, often working at their home.  The rest are identified as corresponding through government accounts.

One might ask what is happening around this time period?  In a word, Benghazi.  Five days after the assault, The U.S. Ambassador to the United Nations, Susan Rice, made a series of television appearances explaining that the attack on the U.S. Benghazi mission was a spontaneous response a  ``hateful video''  which had hours earlier caused violent protests in Cairo against the U.S. Embassy \cite{faceTheNationSept16}.   Four years later, it will be revealed as part of the House Select Committee on Benghazi report \cite{BenghaziReport} that Secretary Clinton had as early as September 12, 2012, communicated to Egyptian Prime Minister Kandil that the attack was known to be planned by Al Qaeda affiliates, it was not a protest, and that it had nothing to do with the film \cite{ClintonAdmitsAttackSept12}.  By the end of September the U.S. administration was calling it a terrorist attack \cite{factCheckerBenghazi}.  A month after the attack the administration was defending itself against charges that it deliberately downplayed a terrorist attack in Libya for political reasons in a presidential election year \cite{susanRiceWashPost2012}.   In October, focus shifted to the security at Benghazi and whether it had been reduced prior to the attack  \cite{cbsSecurity}.  The House Committee on Oversight \& Government Reform began looking into the security aspect \cite{cbsNewsHouseOversight2012} in earnest and found evidence that security issues were known and that requests from the U.S. Embassy in Libya for additional security personnel had been turned down by the State Department \cite{BenghaziSecurityTestimony}.  On October 26, CIA operatives who had defended the Benghazi mission against the terrorist attacks came forward with their story of what went on, on the ground.  They maintain that it was an organized attack and tell reporters that they were told to ``stand down'' \cite{standDownOrder}.  From October 26 to November 9 2012, no email appears in the archive as being sent from Secretary Clinton.

@@

To explore this possibility,  the time slider was adjusted to the relatively small span of thirty days.  This would allow us to detect gaps in email of smaller sizes.  To focus attention on Secretary Clinton's behaviour, only those emails sent from Secretary Clinton were displayed.  
Moving the slider, from left to right, from early 2009 to early 2013, would allow us to move a one month wide magnifying glass across the whole of the archive.  A few contiguous days with no email from Clinton should appear visually as a small horizontal line at zero in the email volume (magenta coloured and lower) time series.  Any such horizontal patch will be identified by its date.

We found ten more patches, in addition to the original obvious email gap of Section~\ref{sect:emailGap}.  So as to get some idea of the context of the patch, and to illustrate the advantage of coupling a visual analytic tool with an internet search engine, for every patch we found we performed a coarse internet search to see what might turn up, if anything to explain it.  If, for example, we found a patch of no emails from Clinton in March 2009, then a Google search as shown in Figure~\ref{fig:googleSearch}
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.75\linewidth]{googleSearch}
\caption{A coarse Google search}
\label{fig:googleSearch}
\end{center}
\end{figure}
would be tried to see what might be found in the news that could relate to that date.

\subsubsection{You have no mail \ldots}
The first, and largest, patch occurs at the very beginning of Clinton's tenure as Secretary of State.  It is  obvious even from the full time series of Figure~\ref{fig:VolumeAll}.  Essentially no email from Secretary Clinton, and little from anyone else appears until mid to late April 2009.

In time order, the patches and what we learned from each Google search was as follows.

We found ten more places in the archive which we identified as gaps to be explored.

where there were zero emails from Clinton

Dates:
\begin{itemize}
\item None until mid April 2009
\item June 8 or 9, June 15 to about the 20th, 2009
\item July 17-20,  around 27, 2009 
\item Day or 2 around Oct 13, 14, 2009
\item April 10-15 or so? 2011
\item one or two days late August, 2011. (no mail)
\item end of march first half of April 2012
\item August 27-30  2012 
\item low first week of September, 2012
\item 2 weeks end of october, up to Nov 10 approx 2012
\item Dec 7- 17? 2012,  Again about Dec 25 2012 to Jan 1, 2013
\end{itemize}

Contentious issues:
\begin{itemize}
\item private email server
\item classified documents (outside of state)
\item Sidney Blumenthal
\item Benghazi spin handling of media (Susan Rice)

\item scrubbing of her server (bleachbit)
\item missing email from online email State department
\end{itemize}

Learn:
\begin{itemize}
\item inner circle over time (state or not)
\item spike in email around Libyan revolution
\item gap in the email
\item daily email patterns, server behaviour (daylight savings time)
\end{itemize}

 Filter:
\begin{itemize}
\item time
\item redacted or not
\item FOIA exemption codes
\end{itemize}

 Content:
\begin{itemize}
\item Term frequency, TFIDF
\end{itemize}

>>>>

\abstract{We present a web-based visualization that allows the user to interactively filter and display characteristics of 32,795 of Hillary Clinton's emails as provided by Wikileaks.

The visualization focuses on the meta-data of each email, including its senders, receivers, and the timestamp the email appeared on the Clinton server (from the Wikileaks source).  An interactive time range slider filters all email and all displays automatically update to changes in the slider.  The main display shows Clinton's most frequent correspondents arranged as nodes of a spoked graph with Clinton at the centre.  Volume determines the thickness of each spoke and high volume determines an inner circle whose spokes are shortened.  Correspondents and their edges are coloured according to whether that email account could be identified as being an approved Federal government account or not.  A second display shows two daily time series: the total number of emails for that day, and the number meeting selection criteria.  A third display shows a scatterplot of the time of day versus the day on which that email appeared.  Scatterplot points are coloured by whether the email was redacted or not.

@@

The paper is organized as follows.  We provide a brief timeline on the private email server to provide sufficient context for the work.  Readers more familiar with the background may wish to skip to the next three sections where the implementation is described.  Section \ref{sect:data} discusses the source of the data, its extraction and cleaning,  details of the metadata,  redaction codes as quasi-metadata,  and our limited use of the actual email content. 
Section \ref{sect:Displays} describes the displays used and illustrates them on the data. These include a spoked network plot, an email volume time series plot, a scatterplot of daily sending times, a barplot of exemption codes, and term frequency displays. Section \ref{sect:Filters} presents the interactive filters that can be applied to change the displays.  With the implementation described, some interactive analyses of the Clinton email archive are conducted to give a quick overview of the utility of the service.  We find some interesting, and potentially contentious, patterns in the email correspondence.  The section closes with  caveats on over interpretation of these analyses.  Some closing thoughts are given in the final section.

%\section{Background}
%\label{sect:Background}
%
%\subsection{A brief timeline on the private email server}
%\label{sect:Background:Timeline}
\section{A brief timeline on the private email server}
\label{sect:Timeline}
While the following timeline is somewhat abbreviated, it should serve to raise some of the major issues and concerns related to Clinton's private email server and its contents.  It also introduces some of the principal characters involved.  More complete and in depth timelines are readily available elsewhere (e.g. \cite{attkissonTimeline, thompsonTimeline, WashPostTimeline, clintonWikipedia, TimeMagEverything}).

@@

Public knowledge that a private email server was being used by Secretary Clinton and others for State Department and personal communications did not surface until March 2015 \cite{NewYorkTimes2015} during the course of a U.S. Congressional investigation \cite{BenghaziReport} of the September 11, 2012 attack by militants on U.S. compounds in Benghazi Libya.

@@

Combetta, Mills, and Samuelson were later granted partial immunity by the Justice Department during the FBI investigations into the private email server, as were two others: Bryan Pagliano (original server manager) and  John Bentel (former director of Information Resources Management for the State Department's Executive Secretariat) \cite{immunityPolitico, immunityDailyCaller, immunityIT}.

On April 12, 2015 Clinton announces that she is running for the U.S. Presidency.  On July 24, 2015, inspectors general for the State Department  and the national intelligence agencies announce that they found classified information in the emails and that the information was classified at the time sent \cite{serverClassified}, though her campaign declared that they must have been classified after the fact. On August 19, 2015, Clinton calls the allegation of mishandling classified information a ``disagreement between agencies'' \cite{clintonDenialGuardian}.

Nearly one year later, July 5, 2016,  FBI Director James Comey recommended that no charges be laid against Clinton for her use of a private email server \cite{nochargeFBI}.  On October 28, 2016, Comey revealed that in a separate investigation into former Congressman Anthony Weiner, that emails belonging to his wife Huma Abedin had been found on his laptop.  Since Abedin was a close aid to former Secretary Clinton, FBI investigations were reopened into the private server usage but closed again by Comey on November 6, 2016 without charges being laid \cite{nochargeFBINov}.  In both cases, Comey and the FBI are criticized by pundits from different political parties.

@@

\subsection{Metadata}
\label{sect:data:metadata}
From the HTML of each email we extract as best we can the identity of the persons sending and receiving the email as well as the date and time at which the email was processed by the server. The fields used were the address to, address from, contact name to, contact name from, subject line, and time. As well, forwarding chains of email addresses were captured through the identification of any to or from fields followed by emails within the text. In cases where no contact name was present the address was substituted. When the address was missing no imputation was completed. Carbon copy information was not extracted.

The HTML was constructed from email printed out, redacted, and then provided as pdfs by the State Department.  Consequently, detailed email header information as would normally be available electronically is mostly missing.   All time stamps appear to be the local date and time at which the server sent or received that email (e.g. no time zone or other source time or IP chain information is available).  Moreover, because of redaction (typically FOIA exemption B6 ~\cite{FOIA}),  sender and receiver emails may have truncated domains,  contain only the person's name, or be missing altogether.  In cases where both \texttt{From} and \texttt{To} are entirely missing we do not impute values  (e.g. see \texttt{https://wikileaks.org/clinton-emails/emailid/31599}).

@@

\section{Displays}
\label{sect:Displays}
\subsection{Inner circle}
\label{sect:Displays:circle} 
This display takes the most frequent correspondents (max. 20) in the selected emails and arranges their email addresses  at the ends of  equiangular spokes having Secretary Clinton as hub.  The hub is coloured red to show that this account is known not to be a government sponsored account (either \texttt{.gov} of \texttt{.mil}).  Every account that is identifiably not a government sponsored account is coloured red.  Those that are identifiably government are coloured either blue (for \texttt{.gov}) or black (for \texttt{.mil}, if any appear).  Those which can not be determined as either government or not, are coloured orange.

\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{SpiralNetworkFullTime}
\caption{Inner Circle over all correspondence}
\label{fig:InnerCircleAll}
\end{center}
\end{figure}
Figure~\ref{fig:InnerCircleAll} shows the top 20 correspondents (or ``Inner Circle'')  over all emails in the collection.
Note that Huma Abedin appears both as a blue  \texttt{.gov} account and as a red non-government (\texttt{clintonemail.com}) account.   The other red account is that of Sidney Blumenthal.  This has been a source of some controversy  \cite{BlumenthalControversy, NYTBlumenthalBenghazi}, since he had been rejected by the State Department yet appears to provide Clinton advice throughout her tenure.  Wendy Sherman appears as orange. Sherman was appointed by Clinton to be Under Secretary of State for Political Affairs in 2011 and the high frequency of Sherman emails dates from this time.  Nevertheless, from this archive the email address could not be identified definitively as being either government or non-government.  Either the address was redacted for privacy reasons (a B6 FOIA exemption), or it was simply unavailable in the header provided \cite{WikileaksPull}.

The width of each spoke is proportional to volume of emails between correspondents, the wider is their connecting spoke.  Whenever the difference in volume is great enough,  correspondents are separated into two groups: those with the greatest correspondence have shorter spokes visually placing them ``closer'' to Clinton.  As Figure~\ref{fig:InnerCircleAll} shows shows, Clinton's closest inner circle are her closest aides Huma Abedin, Cheryl Mills, and Jake Sullivan.
\subsection{Email volume}
\label{sect:Displays:volume}
Figure~\ref{fig:VolumeAll} 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{VolumeFullTime}
\caption{Email Volume: All emails (top) ; Those sent by Clinton (bottom)}
\label{fig:VolumeAll}
\end{center}
\end{figure}
presents two time series showing the volume of emails for each day.  In grey at top is the daily total; in magenta at bottom is the total for the subset chosen by the filters.  In Figure~\ref{fig:VolumeAll}, the bottom series is the total daily emails sent from Clinton to others.

@@

\subsection{Term frequency and TFIDF}
\label{sect:Displays:TFIDF}
The 20 terms (stemmed words excluding stopwords) which appear in the greatest number of the emails selected are displayed as those having ``Highest Frequency''.  
\begin{figure}[h]
\begin{center}
\begin{tabular}{ccc}
\includegraphics[width=0.4\linewidth]{HighFreqAll} &
$~~~~$ &
\includegraphics[width=0.4\linewidth]{HighTFIDFAll} 
\end{tabular}
\caption{Top 20 terms appearing in all emails}
\label{fig:top20all}
\end{center}
\end{figure}
Those terms which appear frequently within some emails but less frequently across emails will have a high tf-idf (term frequency - inverse document frequency) score.  The top 20 scoring of these in the selected emails are shown in the ``TFIDF'' display. 
The top 20 shown in Figure~\ref{fig:top20all} are for all emails in the corpus; these will change depending on the filtering.

@@

\subsection{The last two years}
One might begin an interactive analysis by looking at the email patterns in the last two years of Clinton's tenure as Secretary of State.  
Figure~\ref{fig:lastTwoYears} 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{EmailVolumeB1LastTwoYears}
\caption{Last two years:  Selected Emails are those that are B1 redacted}
\label{fig:lastTwoYears}
\end{center}
\end{figure}
shows the email volume time series over this period. 
This is accomplished by simply having the right end of the time slider of Figure~\ref{fig:timeSlider} as far right as possible and then moving the left most end towards the middle of the time period until January 1, 2011 is reached.   Note, however, that in addition to filtering on time we have chosen to also filter on FOIA exemption code, selecting only those emails containing one or more B1 (national security and foreign policy matters) redactions.  The magenta time series at the bottom then shows all emails, sent or received by Clinton, in the last two years of her tenure which were redacted in part by the State Department (after the fact, not at the time) as being B1 FOIA exempt.  As can be seen, there are relatively few of these.

Recall from Section~\ref{sect:Displays:circle} that the tallest spike of Figure~\ref{fig:lastTwoYears} corresponds to the beginning of the battle for Tripoli (August 21, 2011).   The next largest spikes appear at the far left of this time window, at the beginning of 2011.  We might choose to compress the time window more, to focus on this section of the emails.

@@

Slightly to the right of this peak, there are several more.   At this time, international forces are gathering to consider Libya and its leader Muammar Gaddafi.  Consulting Jake Sullivan's ``tick tock on Libya'' timeline describing Secretary Clinton's leadership on the Libyan crisis, we move the sliders in to focus on the time range from February 25 when Secretary Clinton announced the suspension of the Libyan embassy in Washington, to March 18, the day after Secretary Clinton had secured ``\ldots Russian abstention and Portuguese and African support for UNSC 1973, ensuring that  it passes. 1973 authorizes a no-fly zone over Libya and `all necessary measures' - code for military action - to protect civilians against Gaddafi's army \ldots'' \cite{tickTockLibya}.  On March 19, 2011 NATO military operations in Libya began \cite{LibyaTimelineWiki}.

The readjusted time series are shown in Figure~\ref{fig:EmailVolumeB1LibyaBuildup}.
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{EmailVolumeB1LibyaBuildup}
\caption{Feb. 25 -- March 15, 2011:  Selected Emails are those that are B1 redacted}
\label{fig:EmailVolumeB1LibyaBuildup}
\end{center}
\end{figure}
The left most spike is the rightmost spike of approximately 80 emails from Figure~\ref{fig:early2011B1}.   
Of course all other displays have been updating every time the time slider is adjusted.

@@

Other restrictions, as measured by the FOIA exemption codes, can be seen in Figure~\ref{fig:FOIAB12011LibyaBuildup}.
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{FOIAB12011LibyaBuildup}
\caption{FOIA code distribution: All  B1 redacted, Feb. 25 -- March 18, 2011}
\label{fig:FOIAB12011LibyaBuildup}
\end{center}
\end{figure}
There appear to be about 100 emails in the selected set.  In addition to having B1 redactions, many of them also have B5 (inter- or intra-agency) and B6 (personal) redactions.

We might now consider the composition of the inner circle for this set of emails.  Figure~\ref{fig:InnerCircle2011B1LibyaBuildup} 
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{InnerCircle2011B1LibyaBuildup}
\caption{Inner Circle: All  B1 redacted, Jan. 1 -- Apr 15, 2011}
\label{fig:InnerCircle2011B1LibyaBuildup}
\end{center}
\end{figure}
shows the set of correspondents.  Not surprisingly, Jake Sullivan is the principal correspondent.  Huma Abedin is next, though she is using two emails, one a government account the other a personal \texttt{clintonemail.com} account.   Sidney Blumenthal, a non-government employee, is in this inner circle of B1 exempted emails.  So too is former U.K. Prime Minister Tony Blair; this is the unidentifiable domain account \texttt{aclb}.
It should be noted that there could be other accounts which do not appear in the inner circle because their emails were not present in the email record.  See Section~\ref{sect:data:metadata}.

@@

The inner circle for all emails from September 11 to November 23, 2012 is shown in Figure~\ref{fig:teaser}(b).  Those closest to Clinton in terms of volume of correspondence are Clinton aides Jake Sullivan, Cheryl Mills, and Monica Hanley.  Monica Hanley also appears with an account whose identity as government or not could not be determined.  Tony Blair and Sidney Blumenthal both appear in the inner circle, as does Wendy Sherman (though the account could not be identified as government or not).  Oscar Flores is an assistant to the Clintons, often working at their home.  The rest are identified as corresponding through government accounts.

One might ask what is happening around this time period?  In a word, Benghazi.  Five days after the assault, the U.S. Ambassador to the United Nations, Susan Rice, made a series of television appearances explaining that the attack on the U.S. Benghazi mission was a spontaneous response a  ``hateful video''  which had hours earlier caused violent protests in Cairo against the U.S. Embassy \cite{faceTheNationSept16}.   Four years later, it will be revealed as part of the House Select Committee on Benghazi report \cite{BenghaziReport} that Secretary Clinton had as early as September 12, 2012, communicated to Egyptian Prime Minister Kandil that the attack was known to be planned by Al Qaeda affiliates, it was not a protest, and that it had nothing to do with the film \cite{ClintonAdmitsAttackSept12}.  By the end of September the U.S. administration was calling it a terrorist attack \cite{factCheckerBenghazi}.  A month after the attack the administration was defending itself against charges that it deliberately downplayed a terrorist attack in Libya for political reasons in a presidential election year \cite{susanRiceWashPost2012}.   In October, focus shifted to the security at Benghazi and whether it had been reduced prior to the attack  \cite{cbsSecurity}.  The House Committee on Oversight \& Government Reform began looking into the security aspect \cite{cbsNewsHouseOversight2012} in earnest and found evidence that security issues were known and that requests from the U.S. Embassy in Libya for additional security personnel had been turned down by the State Department \cite{BenghaziSecurityTestimony}.  On October 26, CIA operatives who had defended the Benghazi mission against the terrorist attacks came forward with their story of what happened, on the ground.  They maintain that it was an organized attack and tell reporters that they were told to ``stand down'' \cite{standDownOrder}.  From October 26 to November 9 2012, no email appears in the archive as being sent from Secretary Clinton.

@@

To explore this possibility,  the time slider can be adjusted to a relatively small span, say of thirty days.  This would allow gaps in email of smaller sizes to be detected.  To focus attention on Secretary Clinton's behaviour, only those emails sent from Secretary Clinton would be displayed.  
Moving the slider, from left to right, from early 2009 to early 2013, allows one to move a one month wide magnifying glass across the whole of the archive.  A few contiguous days with no email from Clinton will appear as a small horizontal line at zero in the email volume (magenta coloured and lower) time series.  Once found, the sliders can be further narrowed to identify the dates of the patch.

Following this approach, ten more patches can be found (in addition to that of Section~\ref{sect:emailGap}).  With the dates in hand, an internet search can be conducted to see what, if anything, might turn up to explain the patch.
\subsubsection{You have no mail \ldots}
\label{sect:youHaveNoMail}
The first, and largest, patch occurs at the very beginning of Clinton's tenure as Secretary of State.  It is  obvious even from the full time series of Figure~\ref{fig:VolumeAll}.  Essentially no email from Secretary Clinton, and little from anyone else appears until mid to late April 2009.

This is the time when Secretary Clinton's email habits first run afoul of the State Department as Clinton insists on using her personal Blackberry for all of her email communications.  The issue came to a head in February when National Security officials and intelligence specialists explain the risks to Cheryl Mills.  The officials were apparently not aware  of the private email server (e.g. see \cite{TakingRootWashPost}).  
In time order, the patches and what we learned from each Google search was as follows.

Three of the patches appear to be easily interpreted.   June 17 to June 23, 2009 coincides with Clinton's surgery for an elbow fracture after a fall \cite{ClintonFracture2009}.  A month later, from July 17 to July 23 another patch appears.  This coincides with Secretary Clinton's State visits to India and Thailand \cite{ForeignSched}.  Similarly, the absence of email from Clinton from October 11 to  October 13, 2009 corresponds to a State visit to Russia as shown in  Figure~\ref{fig:ClintonZerosOctober2009AndForeignSchedule} \cite{ForeignSched, visitRussia}.
\begin{figure}[h]
\begin{center}
\includegraphics[width=0.95\linewidth]{ClintonZerosOctober2009AndForeignSchedule}
\caption{Oct 3 -- Oct 27, 2009:  Selected Emails are from Clinton; Scheduled foreign travel marked in blue.}
\label{fig:ClintonZerosOctober2009AndForeignSchedule}
\end{center}
\end{figure}

The next patch chronologically comes in 2011, from April 11 to April 15.   On the last two days of the patch, Secretary Clinton met with other NATO Foreign Ministers in Germany \cite{ForeignSched}.   Notably, in an April 10, 2011 email, Huma Abedin forwards Clinton the ``Stevens Update'' from Timmy Davis.  The email describes Ambassador Stevens's concerns about the situation in Ajdabiyah (less than 150 km from Benghazi) where he is considering leaving Benghazi \cite{StevensUpdate}.
